Extraction of programs from proofs

### 2022-10-03 slides-fischbachau2022.tex
More notes

    The methods presented in the second and in the third lecture are deeply
    related to the \emph{dynamical approach to algebra} reported on in Stefan
    Neuwirth's course. Coherent (and geometric) logic as in Marc Bezem's course
    also plays an important role in this toolbox. It is greatly informed by a
    categorical analysis as provided in Steve Awodey's course, particularly so
    for the third lecture. The first and the second lecture overlap with
    Chuangjie Xu's course, particularly in employing variants of the
    double-negation translation.

    This course is set in an informal constructive metatheory. Formalization
    would both be possible in type theory as in Fredrik Nordvall-Forsberg's
    course or in constructive set theory as in Hajime Ishihare's course.
    All invocations of the transfinite such as by the law of excluded middle or
    by the axiom of choice will be explicitly pointed out. For primers to
    constructive mathematics, enjoy
    \fixedhref{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej
    Bauer's 2013 IAS talk}, its \fixedhref{xxx}{written version} or
    \fixedhref{xxx}{these course notes}.
  }

@@

    It is astonishing what she started and which long way we have come!

@@

  \jnote{1}{
    For a written primer to realizability theory, see
    \fixedhref{http://math.andrej.com/asset/data/c2c.pdf}{Andrej Bauer's course
    notes} and the notes by
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~streicher/REAL/REAL.pdf}{Thomas
    Streicher}.
  }
\end{frame}

@@

    The soundness theorem is proven by an instructive induction on the
    structure of~HA-proofs, verifying that if HA proves a sequent~$\varphi
    \seq{\vec x} \psi$, then there is a realizer for~$\forall x_1\_ \ldots
    \forall x_n\_ (\varphi \Rightarrow \psi)$. The core idea of the proof is to
    verify that every axiom and every rule of~HA is realized. In this way,
    computational content of every axiom and every rule is explicated.
  }

@@

@@

    The multiverse of models of constructive mathematics can be explored from
    the point of view of any base model, and from the point of view of the
    realizability model it looks quite different that from the point of view of
    classical mathematics.
  }

@@

  \jnote{3-4}{
    To set the stage, we analyze the lemma used in the first proof. The
    statement ``every inhabited set of natural numbers contains a minimal
    element'' is not realizable and hence does not directly have
    computational content. In fact, this statement is equivalent to the law of
    excluded middle. Correspondingly, it seems that proofs using it cannot ever
    be constructivized. However, as we will see, much more important for
    extraction of constructive content is the form of the asserted claims than
    the form of the used auxiliary lemmas.
  }

  \jnote{5-}{
    A set~$X$ of natural numbers is \emph{detachable} if and only if for every
    number~$n \in \NN$, either~$n \in X$ or~$n \not\in X$. In classical
    mathematics, every set of natural numbers is detachable. Constructively,
    detachability is a nontrivial and computationally meaningful condition. A
    realizer for detachability of a set~$X$ is a machine which reads a
    number~$n$ as input and determines whether~$n$ belongs to~$X$ or not
    (together with corresponding realizers).
  }

@@

\begin{frame}{The double-negation embedding}
  \jnote{1-3}{
    The starting point of the double-negation embedding is that, while the law
    of excluded middle is not available in intuitionistic logic, the double
    negation of every instance is: For every formula~$\varphi$, the formula
    \[ \neg\neg(\varphi \vee \neg\varphi) \]
    is an intuitionistic tautology:

@@

%    This observation unlocks the following technique to transcribe a classical
%    proof of a statement to an intuitionistic (and even minimal logic) proof of
%    its translation. Most steps in the classical proof will also be acceptable
%    in intuitionistic logic. We simply mimic these. When the classical proof
%    appeals to~\textsc{lem}, we instead use the appropriate doubly negated
%    instance. We
%
%    The difference between intuitionistic and classical logic is concentrated
%    in their treatment of disjunction and existential quantification.
%    Classically, to verify a disjunction~$\varphi \vee \psi$, it suffices to
%    establish~$\neg(\neg\varphi \wedge \neg\psi)$--the intuitionistically
%    weaker claim that it is not the case that neither~$\varphi$ nor~$\psi$
%    hold; similarly, to verify an existential statement~$\exists x\?X\_
%    \varphi(x)$, it suffices to establish~$\neg\forall x\?X\_
%    \neg\varphi(x)$---the claim that nonexistence is impossible.
  }

  \jnote{3}{
    The three parts of the theorem are each proven by induction, on the
    structure of~$\varphi$ for the first two parts and on the structure of
    derivations for the third part. Carrying this out is an instructive
    exercise!
  }

@@

  % xxx importance of remark

@@

\begin{frame}{Barr's theorem / Friedman's trick / A-translation}
  \textbf{Thm.} Let~$\Gamma$ be a set of geometric sequents over a fixed
  signature. Let~$\sigma$ be a geometric sequent. Then the following are
  equivalent:
  \begin{enumerate}
    \addtocounter{enumi}{-1}
    \item $\sigma$ holds for the \good{generic model} of~$\Gamma$ (in its
    classifying topos).
    \item $\sigma$ is provable from~$\Gamma$ in \good{geometric logic}.
    \item $\sigma$ is provable from~$\Gamma$ in \good{intuitionistic logic}.
    \item $\sigma$ is provable from~$\Gamma$ in \bad{classical logic}.
    \item (Assuming \badbox{\textsc{ac}}) $\sigma$ is provable from~$\Gamma$ in
    \bad{classical logic with~\textsc{ac}}.
  \end{enumerate}

@@

\begin{frame}
  \centering\bigskip
  \includegraphics[height=9.5em]{phantoms}
  \medskip

@@

  \visible<3>{\emph{* higher-order proofs of first-order statements using the assumption \\
  that a given (perhaps uncountable) set is countable}}

>>>>

    The methods presented in the second and in the third lecture are deeply
    related to the \emph{dynamical approach to algebra} reported on in Stefan
    Neuwirth's course. Coherent (and geometric) logic as in Marc Bezem's course
    also plays an important role in this toolbox. It is greatly informed by a
    categorical analysis as provided in Steve Awodey's course, particularly so
    for the third lecture. The first and the second lecture overlap with
    Chuangjie Xu's course, particularly regarding the
    double-negation translation.

    This course is set in an informal constructive metatheory. Formalization
    would both be possible in type theory as in Fredrik Nordvall-Forsberg's
    course or in constructive set theory as in Hajime Ishihare's course.
    All appeals to the transfinite such as by the law of excluded middle or
    by the axiom of choice will be explicitly pointed out. For primers to
    constructive mathematics, enjoy
    \fixedhref{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej
    Bauer's 2013 IAS talk}, its \fixedhref{xxx}{written version} or
    \fixedhref{xxx}{these course notes}.
  }

@@

    It is astonishing what she started and what long way we have come!

@@

  \jnote{1}{
    Monika Seisenberger has
    \fixedhref{https://www.proofsociety.org/wp-content/uploads/2018/09/ProgramExtraction_slides.pdf}{many
    and more detailed} slides on this topic.

    For a written primer to realizability theory, see
    \fixedhref{http://math.andrej.com/asset/data/c2c.pdf}{Andrej Bauer's course
    notes} and the notes by
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~streicher/REAL/REAL.pdf}{Thomas
    Streicher}.
  }
\end{frame}

@@

    The soundness theorem is proven by an instructive induction on the
    structure of~HA-proofs, verifying that if HA proves a sequent~$\varphi
    \seq{\vec x} \psi$, then there is a realizer for~$\forall x_1\_ \ldots
    \forall x_n\_ (\varphi \Rightarrow \psi)$. The core idea of the proof is to
    verify that every axiom and every rule of~HA is realized. In this way,
    computational content of every axiom and every rule is explicated.

    For instance, the statement (with no free variables)
    \[ \bot \Rightarrow \varphi \]
    is realized by any number~$e \in \NN$ such that for every~$r \in \NN$
    with~$r \realizes \varphi$ (this condition is never satisfied), $e \cdot r
    \downarrow$ and~$e \cdot r \realizes \varphi$, so by any number whatsoever.
  }

@@

  \jnote{8}{
    By \emph{countable choice}, we mean the statement
    \[ (\forall x\?N\_ \exists y\?A\_ \varphi(x,y)) \Longrightarrow
      (\exists f\?A^N\_ \forall x\?N\_ \varphi(x,f(x))). \]
    Up to some repackaging, this statement is realized by the identity machine
    which simply outputs its input unchanged.

    Choice for higher type fails in the realizability model. For instance, the
    statement
    \[ (\forall f\?N^N\_ \exists y\?A\_ \varphi(f,y)) \Longrightarrow
      (\exists \theta\?A^{N^N}\_ \forall f\?N^N\_ \varphi(f,\theta(f))) \]
    is not realized. A realizer for the antecedent would be a machine which,
    given an index for a Turing machine computing a total function~$f : \NN \to
    \NN$, produces a code for a suitable element~$y$. However, this element~$y$
    might not only depend on the extensional input/output behavior of~$f$, but
    also on the specific index (source code), hence wouldn't describe an actual
    function on the set of computable functions~$\NN \to \NN$.

    This issue does not arise with countable choice, as natural numbers have
    unique codes.
  }

  \jnote{9}{
    Similar to Statement~5, Statement~8 can only be formulated in an extension
    of the formal language used here.

    It expresses that, up to unique isomorphism, there is just one model of
    Heyting arithmetic, namely the standard model. This is in stark contrast
    with the situation in classical mathematics, where Gödel's completeness
    theorem/Henkin term models can be used to concoct a host of nonstandard
    models.

    As a consequence, Peano arithmetic is ``quasi-inconsistent'' from the point
    of view of the realizability model, as it is consistent (being
    equiconsistent with HA) but does not admit a model (every model of PA is
    also a model of HA, but HA only has one model, and this does not validate
    the PA-theorem ``every Turing machine terminates or does not terminate'').

    Pointers to relevant literature are in
    \fixedhref{https://arxiv.org/pdf/2204.00948.pdf}{this survey paper}
    (Example~8 there). Also see
    \fixedhref{https://www.ps.uni-saarland.de/Publications/documents/HermesKirst_2022_An-Analysis.pdf}{the
    2022 paper by Marc Hermes and Dominik Kirst}, particularly also the final
    paragraph of their Section~8.1 which alludes to a result in a different
    direction.
  }

@@

    The multiverse of models of constructive mathematics can be explored from
    the point of view of any base model, and from the point of view of the
    realizability model it looks quite different than from the point of view of
    classical mathematics.
  }

@@

  \jnote{3-4}{
    To set the stage, we analyze the lemma used in the first proof. The
    statement ``every inhabited set of natural numbers contains a minimal
    element'' is not realizable and hence does not directly have
    computational content. In fact, this statement is equivalent to the law of
    excluded middle. Correspondingly, it seems that proofs using it cannot ever
    be constructivized. However, as we will see, much more important for
    extraction of constructive content is the form of the asserted end results than
    the form of auxiliary lemmas used in a proof.
  }

  \jnote{5-}{
    A set~$X$ of natural numbers is \emph{detachable} if and only if for every
    number~$n \in \NN$, either~$n \in X$ or~$n \not\in X$. In classical
    mathematics, every set of natural numbers is detachable. Constructively,
    detachability is a nontrivial and computationally meaningful condition. A
    realizer for detachability of a set~$X$ is a machine which reads a
    number~$n$ as input and determines whether~$n$ belongs to~$X$ or not
    (outputting also corresponding realizers).
  }

@@

\begin{frame}{The double-negation embedding}
  \jnote{1-3}{
    Minimal logic is intuitionistic logic minus \emph{ex falsum quodlibet}
    ($\bot \seq{\vec x} \varphi$).

    The double-negation embedding builds on a fundamental
    observation: While the law
    of excluded middle is not available in minimal or intuitionistic logic, the double
    negation of every instance is---for every formula~$\varphi$, the formula
    $\neg\neg(\varphi \vee \neg\varphi)$
    is an intuitionistic tautology:

@@

    \vspace*{-0.6em}
    Also, a doubly negated statement is not a dead end. Instead, we can
    continue reasoning, by the following tautology (``Kleisli extension'',
    ``monadic bind'', ``nucleus axiom''):

    \vspace*{-1.6em}
    \[ (\neg\neg\varphi \wedge (\varphi \Rightarrow \neg\neg\psi))
    \Longrightarrow \neg\neg\psi \]
  }

  \jnote{3}{
    \vspace*{-0.8em}
    The three parts of the theorem are each proven by induction, on the
    structure of~$\varphi$ for the first two parts and on the structure of
    derivations for the third part. Carrying this out is an instructive
    exercise!
  }

@@

  \jnote{5}{
    Live demo with
    \fixedhref{https://agdapad.quasicoherent.io/~AgdaPadova/html/Dickson-pc2022.html}{Agda
    code for the theorem on all infinite sequences of natural numbers being
    good}:

    \begin{enumerate}
    \item\justifying The first proof there implements the classical proof,
    postulating the law of excluded middle as an unprovable axiom. Thereby Agda
    can verify the classical proof to be correct, however trying to run the
    proof will get stuck on the postulated \textsc{lem}-oracle.

    \item The second proof rewrites the statement of the theorem and its proof
    by the double-negation translation. The result is a constructive proof
    which doesn't rely on postulates; however there is still no direct
    computational content, as the asserted claim is a (doubly) negated
    statement.

    \item In a logical sleight of hand, the third proof imports the
    second proof but specializes~$\bot$, which was an arbitrary constant there,
    to the asserted claim. The tautology~$\negg\negg\bott \Rightarrow \bott$ of
    minimal logic is then used to escape the double-negation monad and obtain a
    constructive proof of the full result.
    \end{enumerate}
  }

@@

\begin{frame}{Barr's theorem / Friedman's trick / A-translation}
  \jnote{4}{
    The proof transformation for~``$3 \Rightarrow 2$'' is explicit in
    nature, feasible in practice and increases the proof length only
    polynomially. There is also an
    \fixedhref{https://drops.dagstuhl.de/opus/volltexte/2022/16776/pdf/LIPIcs-TYPES-2021-7.pdf}{intriguing
    alternative approach by Giulio Fellin, Sara Negri and Eugenio Orlandelli}
    employing cut elimination (thereby increasing the proof length
    substantially, but conceptually beneficial in other aspects). Furthermore,
    this metatheorem can also be cast in semantic terms:

    Every Grothendieck topos admits a cover (a surjective geometric morphism from
    another Grothendieck topos) by a Grothendieck topos which is boolean. This
    can be obtained by first adjoining a \emph{generic proposition}~$\chi$ and then
    taking sheaves with respect to the modality~$(({\cdot} \Rightarrow \chi)
    \Rightarrow \chi)$.

    Applied to the classifying topos of~$\Gamma$, this insight yields a proof
    of~``$3 \Rightarrow 0$'', since pullback along surjective geometric
    morphisms reflects validity of geometric sequents.

    Furthermore, every Grothendieck topos admits a cover by a Grothendieck
    topos over a complete Boolean algebra. Assuming Zorn's lemma in the
    metatheory, this topos validates~\textsc{lem} and~\textsc{zorn},
    hence~\textsc{ac}. This shows~``$4 \Rightarrow 0$''.

    An introduction to topos-theoretic generic models with a view towards
    applications in constructive commutative algebra is contained in
    \fixedhref{https://arxiv.org/abs/2012.13850}{this survey}.
    \vspace*{2em}
  }

  \textbf{Thm.} Let~$\Gamma$ be a set of geometric sequents over a fixed
  signature. Let~$\sigma$ be a geometric sequent. Then the following are
  equivalent:
  \begin{enumerate}
    \addtocounter{enumi}{-1}
    \item $\sigma$ holds for the \good{generic model} of~$\Gamma$ (in its
    classifying topos).
    \item $\sigma$ is provable from~$\Gamma$ in \good{geometric logic}.
    \item $\sigma$ is provable from~$\Gamma$ in \good{intuitionistic logic}.
    \item $\sigma$ is provable from~$\Gamma$ in \bad{classical logic}.
    \item (Assuming \badbox{\textsc{zorn}}) $\sigma$ is provable from~$\Gamma$ in
    \bad{classical logic with~\textsc{ac}}.
  \end{enumerate}

@@

\begin{frame}
  \jnote{3}{
    \textbf{Thm. 1.} From every proof~$p$ of a statement~$\varphi$ such that
    \vspace*{-0.5em}
    \begin{enumerate}
      \item the assertion~$\varphi$ is a first-order statement,
      \item the proof~$p$ is constructive,
      \item the proof~$p$ is formulated in a certain higher-order system (in
      particular, the proof may, unlike its end result, freely employ
      higher-order notions) and
      \item for a certain set~$X$ appearing in~$p$, the proof assumes that~$X$
      is countable,
    \end{enumerate}
    \vspace*{-0.5em}
    a constructive proof of the same statement~$\varphi$, set in the same
    system, but \emph{without requiring the countability assumption} can be
    extracted.
  }

  \jnote{4}{
    \textbf{Thm. 2.} From every proof~$p$ of a statement~$\varphi$ such that
    \vspace*{-0.5em}
    \begin{enumerate}
      \item the assertion~$\varphi$ is a geometric implication,
      \item the proof~$p$ is classical (constructive + \textsc{lem}, but still no
      \textsc{zorn}),
      \item the proof~$p$ is formulated in a certain higher-order system (in
      particular, the proof may, unlike its end result, freely employ
      higher-order notions) and
      \item for a certain set~$X$ appearing in~$p$, the proof assumes that~$X$
      is countable,
    \end{enumerate}
    \vspace*{-0.5em}
    a classical proof of the same statement~$\varphi$, set in the same
    system, but \emph{without requiring the countability assumption} can be
    extracted.
  }

  \centering\bigskip
  \includegraphics[height=9.5em]{phantoms}
  \medskip

@@

  \visible<3->{\emph{* higher-order proofs of first-order statements using the assumption \\
  that a given (perhaps uncountable) set is countable}}
